---
title: "Reproducible Research - Course Project 2"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Ajit Kumar
07/01/2021

## Exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database - Health and Economic Impacts

### Synopsis


This is a second course project for Reproducible Research course which is part of the Coursera's Data Science Specialization.  

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.  

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.  

The analysis of the data shows that tornadoes, by far, have the greatest health impact as measured by the number of injuries and fatalities The analysis also shows that floods cause the greatest economic impact as measured by property damage and crop damage.  

### **Data Processing**
### Load Libraries and prepare the R environment

I used these librarys in my analysis:
```{r}
library(ggplot2)  
library(plyr) 
library(dplyr)
```
### Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

[storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)[47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.  

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  

* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.  

### Assignment

The basic goal of this assignment is to explore the NOAA Storm Database and answer the following basic questions about severe weather events.

+ Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
+ Across the United States, which types of events have the greatest economic consequences?

### Loading the data

The data was downloaded from the link above and saved on local computer (in setwd command one can replace loacal file path with path of folder where the data was downloaded). Then it was loaded on the R using the read.csv command. If object strom.data is already loaded, use that cached object insted of loading it each time the Rmd file is knitted.

```{r}
if(!exists("storm.data")) {
    storm.data <- read.csv(bzfile("repdata_data_StormData.csv.bz2"),header = TRUE)
  }
```

### Examine the data set

In storm.data there is 37 columns (variables) and 902,297 rows (records).

```{r}
dim(storm.data)
```

Examine the structure of the data
```{r}
str(storm.data)
```

### Extracting variables of interest for analysis of weather impact on health and economy

From a list of variables in storm.data, these are columns of interest:  

Health variables:  
* FATALITIES: approx. number of deaths  
* INJURIES: approx. number of injuries  

Economic variables:  

* PROPDMG: approx. property damags  
* PROPDMGEXP: the units for property damage value  
* CROPDMG: approx. crop damages  
* CROPDMGEXP: the units for crop damage value  

Events - target variable:  

*  EVTYPE: weather event (Tornados, Wind, Snow, Flood, etc..)  

Extract variables of interest from original data set:  
```{r}
vars <- c( "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
mydata <- storm.data[, vars]
```

Check the last few rows in data set (in firs years of recording there are many missing (NA) values):
```{r}
tail(mydata)
```
###Checking for missing values
In every analysis we must the check number of missing values in variables.  

Check for missing values in health variables - there is no NA's in the data.  

```{r}
sum(is.na(mydata$FATALITIES))
```
```{r}
sum(is.na(mydata$INJURIES))
```
Check for missing values in economic variables for "size" of damage - there is no NA's in the data.
```{r}
sum(is.na(mydata$PROPDMG))
```
```{r}
sum(is.na(mydata$CROPDMG))
```
Check for missing values in economic variables for units damage - there is no NA's in the data.
```{r}
sum(is.na(mydata$PROPDMGEXP))
```
```{r}
sum(is.na(mydata$CROPDMGEXP))
```

### Transforming extracted variables
Listing the first 10 event types that most appear in the data:
```{r}
sort(table(mydata$EVTYPE), decreasing = TRUE)[1:10]
```

We will group events like TUNDERSTORM WIND, TUNDERSTORM WINDS, HIGH WIND, etc. by containing the keyword 'WIND' as one event WIND. And we will transform other types of events in a similar way. New variable EVENTS is the transform variable of EVTYPE that have 10 different types of events: HEAT, FLOOD, etc., and type OTHER for events in which name the keyword is not found.
```{r}
# create a new variable EVENT to transform variable EVTYPE in groups
mydata$EVENT <- "OTHER"
# group by keyword in EVTYPE
mydata$EVENT[grep("HAIL", mydata$EVTYPE, ignore.case = TRUE)] <- "HAIL"
mydata$EVENT[grep("HEAT", mydata$EVTYPE, ignore.case = TRUE)] <- "HEAT"
mydata$EVENT[grep("FLOOD", mydata$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
mydata$EVENT[grep("WIND", mydata$EVTYPE, ignore.case = TRUE)] <- "WIND"
mydata$EVENT[grep("STORM", mydata$EVTYPE, ignore.case = TRUE)] <- "STORM"
mydata$EVENT[grep("SNOW", mydata$EVTYPE, ignore.case = TRUE)] <- "SNOW"
mydata$EVENT[grep("TORNADO", mydata$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
mydata$EVENT[grep("WINTER", mydata$EVTYPE, ignore.case = TRUE)] <- "WINTER"
mydata$EVENT[grep("RAIN", mydata$EVTYPE, ignore.case = TRUE)] <- "RAIN"
# listing the transformed event types 
sort(table(mydata$EVENT), decreasing = TRUE)
```
Checking the values for variables that represent units od dollars:
```{r}
sort(table(mydata$PROPDMGEXP), decreasing = TRUE)[1:10]
```
```{r}
sort(table(mydata$CROPDMGEXP), decreasing = TRUE)[1:10]
```
There is some mess in units, so we transform those variables in one unit (dollar) variable by the following rule:  
* K or k: thousand dollars (10^3)  
* M or m: million dollars (10^6)  
* B or b: billion dollars (10^9)  
* the rest would be consider as dollars  

New variable(s) is product of value of damage and dollar unit.  
```{r}
mydata$PROPDMGEXP <- as.character(mydata$PROPDMGEXP)
mydata$PROPDMGEXP[is.na(mydata$PROPDMGEXP)] <- 0 # NA's considered as dollars
mydata$PROPDMGEXP[!grepl("K|M|B", mydata$PROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
mydata$PROPDMGEXP[grep("K", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "3"
mydata$PROPDMGEXP[grep("M", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "6"
mydata$PROPDMGEXP[grep("B", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "9"
mydata$PROPDMGEXP <- as.numeric(as.character(mydata$PROPDMGEXP))
mydata$property.damage <- mydata$PROPDMG * 10^mydata$PROPDMGEXP

mydata$CROPDMGEXP <- as.character(mydata$CROPDMGEXP)
mydata$CROPDMGEXP[is.na(mydata$CROPDMGEXP)] <- 0 # NA's considered as dollars
mydata$CROPDMGEXP[!grepl("K|M|B", mydata$CROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
mydata$CROPDMGEXP[grep("K", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "3"
mydata$CROPDMGEXP[grep("M", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "6"
mydata$CROPDMGEXP[grep("B", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "9"
mydata$CROPDMGEXP <- as.numeric(as.character(mydata$CROPDMGEXP))
mydata$crop.damage <- mydata$CROPDMG * 10^mydata$CROPDMGEXP
```

Print of first 10 values for property damage (in dollars) that most appear in the data:  
```{r}
sort(table(mydata$property.damage), decreasing = TRUE)[1:10]
```

Print of first 10 values for crop damage (in dollars) that most appear in the data:  
```{r}
sort(table(mydata$crop.damage), decreasing = TRUE)[1:10]
```

## Analysis
### Aggregating events for public health variables
Table of public health problems by event type  
```{r}
# aggregate FATALITIES and INJURIES by type of EVENT
agg.fatalites.and.injuries <- ddply(mydata, .(EVENT), summarize, Total = sum(FATALITIES + INJURIES,  na.rm = TRUE))
agg.fatalites.and.injuries$type <- "fatalities and injuries"
  
# aggregate FATALITIES by type of EVENT
agg.fatalities <- ddply(mydata, .(EVENT), summarize, Total = sum(FATALITIES, na.rm = TRUE))
agg.fatalities$type <- "fatalities"

# aggregate INJURIES by type of EVENT
agg.injuries <- ddply(mydata, .(EVENT), summarize, Total = sum(INJURIES, na.rm = TRUE))
agg.injuries$type <- "injuries"

# combine all
agg.health <- rbind(agg.fatalities, agg.injuries)

health.by.event <- join (agg.fatalities, agg.injuries, by="EVENT", type="inner")
health.by.event
```

### Aggregating events for economic variables
```{r}
# aggregate PropDamage and CropDamage by type of EVENT
agg.propdmg.and.cropdmg <- ddply(mydata, .(EVENT), summarize, Total = sum(property.damage + crop.damage,  na.rm = TRUE))
agg.propdmg.and.cropdmg$type <- "property and crop damage"

# aggregate PropDamage by type of EVENT
agg.prop <- ddply(mydata, .(EVENT), summarize, Total = sum(property.damage, na.rm = TRUE))
agg.prop$type <- "property"

# aggregate INJURIES by type of EVENT
agg.crop <- ddply(mydata, .(EVENT), summarize, Total = sum(crop.damage, na.rm = TRUE))
agg.crop$type <- "crop"

# combine all
agg.economic <- rbind(agg.prop, agg.crop)


economic.by.event <- join (agg.prop, agg.crop, by="EVENT", type="inner")
economic.by.event
```
##Results
### Across the United States, which types of events are most harmful with respect to population health?
```{r}
# transform EVENT to factor variable for health variables
agg.health$EVENT <- as.factor(agg.health$EVENT)

# plot FATALITIES and INJURIES by EVENT
health.plot <- ggplot(agg.health, aes(x = EVENT, y = Total, fill = type)) + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Event Type") + 
  ylab("Total number of health impact") +
  ggtitle("Weather event types impact on public health") +
  theme(plot.title = element_text(hjust = 0.5))
print(health.plot)  
```

The most harmful weather event for health (in number of total fatalites and injuries) is, by far, a tornado.  
### Across the United States, which types of events have the greatest economic consequences?
```{r}
# # transform EVENT to factor variable for economic variables
agg.economic$EVENT <- as.factor(agg.economic$EVENT)

# plot PROPERTY damage and CROP damage by EVENT
economic.plot <- ggplot(agg.economic, aes(x = EVENT, y = Total, fill = type)) + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Event Type") + 
  ylab("Total damage in dollars") +
  ggtitle("Weather event types impact on property and crop damage") +
  theme(plot.title = element_text(hjust = 0.5))
print(economic.plot) 
```

The most devastating weather event with the greatest economic cosequences (to property and crops) is a flood.
